Density-dependent selection and the limits of relative fitness

### 2018-07-31 manuscript.tex
l

In this appendix we derive Eq.~\eqref{eq:master}. Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.

We start by separating the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components
\begin{equation}
\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i,\label{eq:delt_decomp}
\end{equation}
which vary in relative magnitude depending on the propagule densities $l_i$. The first component, $\Delta_u n_i$, accounts for territories where only one focal propagule is present ($x_i=1$ and $x_j=0$ for $j\neq i$; $u$ stands for ``uncontested''). The proportion of territories where this occurs is $l_i e^{-L}$, and so 
\begin{equation}
\Delta_u n_i=Ul_i e^{-L}=m_i e^{-L}.
\end{equation}

The second component, $\Delta_r n_i$, accounts for territories where a single focal propagule is present along with at least one non-focal propagule ($x_i=1$ and $X_i\geq 1$ where $X_i=\sum_{j\neq i} x_j$ is the number of nonfocal propagules; $r$ stands for ``rare''). The number of territories where this occurs is $Up_i(1)P(X_i\geq 1)=m_i e^{-l_i}(1-e^{-(L-l_i)})$. Thus 
\begin{equation}
\Delta_r n_i = m_i e^{-l_i}(1-e^{-(L-l_i)})\left\langle  \frac{c_i}{c_i +\sum_{j\neq i} c_j x_j } \right\rangle_{\tilde{p}},  \label{eq:deltr}
\end{equation}
where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to the probability distribution $\tilde{p}$ of nonfocal propagule abundances $x_j$, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.

The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($x_i\geq 2$; $a$ stands for ``abundant"). Similar to Eq.~\eqref{eq:deltr}, we have 
\begin{equation}
\Delta_a n_i=U(1-(1+l_i)e^{-l_i})\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\label{eq:delta}
\end{equation}
where $\hat{p}$ is the probability distribution of both focal and nonfocal propagule abundances in those territories where at least two focal propagules landed.

To derive Eq.~\eqref{eq:master} we approximate the expectations in Eq.~\eqref{eq:deltr} and Eq.~\eqref{eq:delta} by replacing $x_i$ and the $x_j$ with ``effective'' mean values as follows 
\begin{equation}
\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{q}}}.\label{eq:meanfieldr}
\end{equation}
\begin{equation}
\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\approx  \frac{c_i \langle x_i \rangle_{\hat{q}}}{\sum_j c_j \langle x_j\rangle_{\hat{q}}}.\label{eq:meanfielda}
\end{equation}
Here the effective means $\langle \rangle_{\tilde{q}}$ and $\langle \rangle_{\hat{q}}$ are taken with respect to new distributions $\tilde{q}$ and $\hat{q}$, respectively. In the following subsection we define $\tilde{q}$ and $\hat{q}$ and explain our reasoning for using these distributions to take the effective means.

\subsection*{The effective distributions $\tilde{q}$ and $\hat{q}$}

The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be consistent between rare and common types. To illustrate, suppose that two identical types (same $b$, $c$ and $d$) are present, that the first type is rare and has vanishingly small density $l_1\rightarrow 0$ while the second is common and has high density $l_2\rightarrow \infty$. Since $l_2\rightarrow \infty$, uncontested territories make up a negligible fraction of the total. The rare type's territorial acquisition is entirely due to $\Delta_r n_1$, while the common type's territorial acquisition is entirely due to $\Delta_a n_2$. To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must therefore be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This behavior is clearly pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes. Thus, predicted growth or decline would depend on an arbitrary assignment of rarity.

In general, the above consistency condition is not satisfied when the conditional Poisson distributions $\tilde{p}$ and $\hat{p}$ are used to calculate the effective means. Under the above limits ($l_1\rightarrow 0$, $l_2\rightarrow \infty$), the right hand side of Eq.~\eqref{eq:meanfieldr} tends to $1/(l_2+1)$, and so $\Delta_r n_1/m_1 \rightarrow 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}. Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}} = l_2$ in Eq. \eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \rightarrow 1/L$. Thus, the identical rare type is pathologically predicted to decline in frequency.

The effective distributions $\tilde{q}$ and $\hat{q}$ are devised to avoid this pathology. The idea is to make the approximation that the distribution for the total number of propagules per territory is the same in all territories. This is only an approximation because conditioning on focal propagules being present does change the distribution of $X$ in the corresponding subset of territories (in the above example, the mean propagule density across all territories is $L$, but in the territories responsible for the growth of the rare type we have $\langle X \rangle_{\tilde{p}}=L+1$).

More formally, let ${\mathbf x}$ denote the vector of propagule abundances $(x_1,\ldots,x_G)$ in a given territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ similarly denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\cdots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\cdots p_G(x_G)$. The corresponding total propagule numbers are denoted $X=\sum_j x_j$ and $X_i=X-x_i$. Then, in territories where one focal propagule and at least one non-focal propagule are present, the effective distribution is defined by 
\begin{equation}
\tilde{q}({\mathbf x_i})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x_i}|X_i=X-1),
\end{equation}
where the total number of propagules $X$ follows a Poisson distribution with mean $L$, and $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$. Similarly, in territories where more than one focal propagule is present, the effective distribution is defined by 
\begin{equation}
\hat{q}({\mathbf x})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x}|x_i\geq 2, X).
\end{equation}

>>>>

In this appendix we derive Eq.~\eqref{eq:master}. Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes. We denote the vector of propagule abundances by ${\mathbf x} = (x_1,\ldots,x_G)$ in a given territory, and the analogous vector of nonfocal abundances by ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$. The corresponding total propagule numbers are denoted $X=\sum_j x_j$ and $X_i=X-x_i$.

Similar to the classic lottery model, our approximation involves replacing the $x_i$ with effective mean values. However, as discussed in the text preceding Eq.~\eqref{eq:master}, it is important to treat the $x_i=1$ case separately when allowing for low propagule densities. We thus start by separating the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components
\begin{equation}
\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i.\label{eq:delt_decomp}
\end{equation}
The relative magnitude of these components depends on the propagule densities $l_i$. The first component, $\Delta_u n_i$, accounts for territories where only one focal propagule is present ($x_i=1$ and $x_j=0$ for $j\neq i$; $u$ stands for ``uncontested''). The proportion of territories where this occurs is $l_i e^{-L}$, and so 
\begin{equation}
\Delta_u n_i=Ul_i e^{-L}=m_i e^{-L}.
\end{equation}

The second component, $\Delta_r n_i$, accounts for territories where a single focal propagule is present along with at least one non-focal propagule ($r$ stands for ``rare''). The number of territories where this occurs is $Up_i(1)P(X_i\geq 1)=m_i e^{-l_i}(1-e^{-(L-l_i)})$. Thus 
\begin{equation}
\Delta_r n_i = m_i e^{-l_i}(1-e^{-(L-l_i)})\left\langle  \frac{c_i}{c_i +\sum_{j\neq i} c_j x_j } \right\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)},  \label{eq:deltr}
\end{equation}
where $\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ denotes the expectation with respect to the conditional probability distribution $p({\mathbf x}|x_i=1,X_i\geq 1)$ of propagule abundances in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.

The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($a$ stands for ``abundant"). Similar to Eq.~\eqref{eq:deltr}, we have 
\begin{equation}
\Delta_a n_i=U(1-(1+l_i)e^{-l_i})\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{p({\mathbf x}|x_i\geq 2)}.\label{eq:delta}
\end{equation}

To derive Eq.~\eqref{eq:master} we approximate the expectations in Eq.~\eqref{eq:deltr} and Eq.~\eqref{eq:delta} by replacing $x_i$ and the $x_j$ with ``effective'' mean values as follows 
\begin{equation}
\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_r}.\label{eq:meanfieldr}
\end{equation}
\begin{equation}
\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{p({\mathbf x}|x_i\geq 2)}\approx  \frac{c_i \langle x_i \rangle_a}{\sum_j c_j \langle x_j\rangle_a}.\label{eq:meanfielda}
\end{equation}
Here $\langle \rangle_r$ and $\langle \rangle_a$ are the effective means, which are defined in the following subsection.

\subsection*{The effective means $\langle \rangle_r$ and $\langle \rangle_a$}

The decomposition Eq.~\eqref{eq:delt_decomp} is exact and involves no additional assumptions. However this decomposition complicates our approximation procedure because the separate components in Eq.~\eqref{eq:delt_decomp} must be approximated in a consistent manner.

To illustrate this consistency requirement, suppose that two identical types (same $b$, $c$ and $d$) are present, the first with small density $l_1\ll 1$ and the second with large density $l_2 \gg 1$. In this case, uncontested territories make up a negligible fraction of $U$, the first type's territorial acquisition is almost entirely due to $\Delta_r n_1$, and the second type's territorial acquisition is almost entirely due to $\Delta_a n_2$. For consistency, the approximate per-capita growth rates in \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$. Even small violations of this consistency condition would mean exponential growth of one type relative to the other. This is pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes. Thus, predicted growth or decline would depend on an arbitrary assignment of rarity.

Suppose that we simply used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$. Then, again using the example from the previous paragraph ($l_1\ll 1$, $l_2\gg 1$), the right hand side of Eq.~\eqref{eq:meanfieldr} would be $\approx 1/(l_2+1)$, and so $\Delta_r n_1/m_1 \approx 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}. Similarly, $\sum_j \langle x_j\rangle_a = l_2$ in Eq.~\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \approx 1/l_2$. Thus, the rare type is pathologically predicted to decline in frequency under this approximation. This inconsistency occurs because the expected total numbers of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $. By approximating the expected outcome of lottery competition with effective means, the rare type is effectively assumed to experience more intense lottery competition.

The effective distributions $\hat{p}$ are devised to avoid this pathology by forcing $\langle X \rangle$ to be same in all territories. They are defined as follows. For each propagule number $x_i$, $\hat{p}(x_i)=p$ which is independent of the other number

 by making the approximation that the distribution for $X$ is the same in all territories.

 Then, in territories where one focal propagule and at least one non-focal propagule are present, the effective distribution is defined by 
\begin{equation}
\tilde{q}({\mathbf x_i})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x_i}|X_i=X-1),
\end{equation}
where the total number of propagules $X$ follows a Poisson distribution with mean $L$, and $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$. Similarly, in territories where more than one focal propagule is present, the effective distribution is defined by 
\begin{equation}
\hat{q}({\mathbf x})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x}|x_i\geq 2, X).
\end{equation}
